\chapter{Temporal Abstraction}
\label{ch:abstraction}

\emph{Abstraction} is the process by transforming a concept or an observable phenomenon in to a more succinct form, typically to retain only information which is relevant for a particular purpose.  For the purpose of inference, the goal of temporal abstraction is to provide a concise, high-level description of a sequential process that facilitates description and comparison among sequences, while simultaneously preserving the information contained in the raw data.

Complex data that has many observations over weeks, or even years, it not easily understood.  Even for trained experts, it can be hard to discern meaning from poorly structured, irregular data.  Data ming facilitates the analysis of data that is too large or complex for human review, and temporal mining focuses on data with an explicit temporal component.  In addition to facilitating the processing of data sets for knowledge discovery and mining, abstraction can benefit interpretation and comprehension to humans too.

Historically, the temporal dimension of data was mostly ignored due to the additional resources that processing or collection entailed.  Now, large-scale data analysis had become more efficient, and the amount of digital artifacts in the world is increasing at an exponential rate.  However, most existing longitudinal analysis techniques are designed for primary data sources, or require a canonical form of a time series.   Along with the many new data mining opportunities, are new limiting factors related to the analysis of noisy, incomplete secondary data sets that are repurposed as research tools.

\input{4_1_techniques}
