\section{Application: Grading and Staging Liver Disease}
\subsection{Background}
`Hepatitis' is a Greek word with the root `hepat' meaning liver and the suffix `itis' indicating inflammation.  It is used to describe a class of viruses that are associated with liver inflammation and is characterized by three main types, A, B and C.  Hepatitis A is associated with acute inflammation of hepatocytes (`cyte' is a suffix meaning cell), and types B and C, chronic inflammation.  Individuals with types B and C have the potential risk of developing more severe stages of the disease that may result in permanent scaring of liver tissue, a condition known as cirrhosis, and hepatocarcinoma, a common type of liver cancer.

An indicter of cirrhosis or hepatocarcinoma is fibrosis of the hepatocytes.  A liver biopsy is the gold standard for the prognosis and treatment of hepatitis B and C and provides information that is used monitor disease progression and inform the treatment of hepatitis B and C patients.     Its primary purpose is to determine the health of the liver by measuring the degree of inflammation and stage of fibrosis.  Two common assessment tools are the Metavir test are shown in Table~\ref{metavir} and the Histologic Activity Index (HAI) shown in Table~\ref{hai}.  The Matavir score gives an indication of the amount of scarring, or fibrosis.  HAI represents an activity level, that corresponds with hepatocyte inflammation.

\begin{table}[ht]
\caption{Metavir Fibrosis Scoring}
\label{metavir}
\vskip 0.15in
\begin{center}
\begin{tabular}{p{0.05\linewidth} p{0.6\linewidth}}
\hline
\textbf{Score}	&  \textbf{Assessment} \\
\hline
0         & no scarring\\
1         & minimal scarring\\
2	     & scarring has occurred and extends outside
 the areas in the liver that contains blood vessels\\
3	     & bridging fibrosis is spreading and
 connecting to other areas that contain fibrosis \\
4	     & cirrhosis or advanced scarring of the liver\\
\hline
\end{tabular}
\end{center}
\vskip -0.1in
\end{table}

\begin{table}[ht]
\caption{Histologic Activity Index}
\label{hai}
\vskip 0.15in
\begin{center}
\begin{tabular}{ll}
\hline
\textbf{Score}	&  \textbf{Assessment}\\
\hline
0 & no inflammation  \\
1-4 & minimal inflammation  \\
5-8 & mild inflammation  \\
9-12 & moderate inflammation  \\
13-18 & marked inflammation  \\
\hline
\end{tabular}
\end{center}
\vskip -0.1in
\end{table}

Although it provides important clinical information to guide the care of chronic disease patients, biopsy is invasive, costly and subject to complications.  Biopsy involves extracting a tissue sample of at least $2-3$ cm in length with a 16-gauge needle that is inserted between two of the patients ribs.  The procedure has been association with complications that can be potentially life-threatening.  Also, can cost hundreds of dollars, and is subject to diagnostic error~\cite{Castera2012}.   For these reasons, alternatives for assessing the stage of liver fibrosis are in great demand and the lack of alternative assessment methods has been noted as a major limitation in both management and research in liver diseases.

\subsubsection{Temporal Indicators}
The progression from chronic hepatitis to hepatocarcinoma via liver cirrhosis is mainly observed in hepatitis C patients.  However, the detailed mechanism of this disease trajectory is not well-understood.  Recent work (cite) has demonstrated that temporal patterns associated with hepatitis indicators can provide useful information about liver activity for predicting fibrosis and present a viable option to more invasive testing.  These indicators are recorded as part of a standard liver panel, a collection of tests that are preformed together and provide useful information to assess liver damage, or urinary tests.

\input{7_1_1_method}
\input{7_1_2_results}
\input{7_1_3_discussion}
